A Stopped Negative Binomial Distribution

### 2016-03-23 stat-and-prob-letters/snb-paper.tex
Integrated Michelle's suggestions.

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. If two or more respond then we reject the null
hypothesis that the treatment is not effective and the trial is deemed
successful. If fewer that two respond then the null is not rejected and the 
trial is terminated.

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, as in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the stage. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

Assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables that stops when either a
specified number of success or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

@@

In this work we derive the distribution of the number of enrollees needed
for either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB) and some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
it's relation to other distributions. Section 2 derives the distribution
based on a defined Bernoulli process and gives some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

\begin{corollary} \label{conditional_distribution}
Assume a trial that stops on the $s$th successs or $t$failure as before. 
Furthermore, assume we observer $s'$ successes and $t'$ failures
where $s' < s$ and $t' < t$.
That is, we see the number of success and failures at a time before
reaching an endpoint. Then the number of subsequent enrollements needed 
to reach an endpoint is SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
\begin{proof}
The result follows directly by realizing that, conditioned on $s'$
success and $t'$ failures, there are $s-s'$ successes needed to reach the
success endpoint and $t-t'$ failures needed to reach the failure endpoint.
\end{proof}

@@

Fig.~\ref{fig:snb_bin_compare} shows the case
where $s=2$, $t=12$, and $p=0.2$. The probability masses represented in
black are equal as are the masses in grey. The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

@@

A similar expression can be derived using the same calculation with
the constraint that $x \leq -\log(p)$. The result
follows from the afore mentioned property of the regularized incomplete
beta function.
\end{proof}

@@

The SNB can be used to extend clinical trial methodology
in at least two areas. First, it can be used to construct nested criteria
for early stopping and efficacy. These trials consist of two-stages.
If the failure endpoint is reached in the first stage then the 
trial stops due to futility. Otherwise, the trial enters the second 
stage, which consists of responders from the first stage along with
new enrollees in the second. Thus the design provides
both this provide early testing for futility along with the ability
to make statistically sound decisions sooner than traditional trials.

@@

There are also several avenues for extending the SNB itself. First is the 
development of a stopped negative multinomial distribution allowing
for more than two endpoints. This extension would have clinical trial
applications including dose escalation where the outcomes are response
to a therapy, failure to respond, and an adverse outcome endpoint where
and enrollee is removed from the trial with neither response or non-response
due to side effects.

>>>>

Consider a prototypical Phase II single-arm clinical trial in which 12 patients
are enrolled and treated. If two or more respond then we reject the null
hypothesis that the treatment is not effective and the trial is deemed
successful. If fewer than two respond then the null is not rejected and the 
trial is terminated.

If all 12 patients are enrolled at once, as in the classic
design, then the sample size is 12. However, as in most clinical trials, the
patients are enrolled sequentially, often with one patient's outcome realized
before the next one enters the trial. In the present example, observing two
successful patients allows us to reach one endpoint so the sample required
could be as small as two. Similarly 11
observed treatment failures also ends the trial. This sampling mechanism, in
which the experiment ends as soon as any of the endpoints is reached, is
call {\em curtailed sampling}. Under curtailed sampling the range of the
sample size is between two and 12.

Assume each patient outcome can be modeled as an independent,
identically distributed Bernoulli($p$) random variable. The trial is realized
as a sequence of these random variables that stops when either a
specified number of successes or failures is reached. In the
previous example suppose two successes were reached after enrolling 10
patients (one in the third step and one at the $10^{th}$). The sample
path is illustrated
in Fig.~\ref{fig:kane_viz}. The vertical axis denotes the number of
successful outcomes. The horizontal axis counts the number of patients that
have been enrolled. The horizontal and vertical boundaries represent
endpoints for the trial.

@@

In this work we derive the distribution of the number of enrollees needed
for either $s$ successes or $t$ failures. We refer to this distribution
as the Stopped Negative Binomial (SNB) and some of its characteristics are
summarized in Tab.~\ref{tab:snb}.
The rest of this paper derives these results
and explores properties of the distribution.
The next section introduces our notation and basic results
including the density of the distribution along with a description of
its relation to other distributions. Section 2 derives the distribution
based on a defined Bernoulli process and gives some basic properties.
Section 3 provides a connection to the Binomial tail probability.
Section 4 derives the posterior distribution using a Beta prior.
Section 5 provides a brief discussion on the use of the distribution
in clinical trials along with future avenues for generalization.

@@

\begin{corollary} \label{conditional_distribution}
Assume a trial that stops on the $s^{th}$ success or $t^{th}$ failure as 
before. 
Furthermore, assume we observer $s'$ successes and $t'$ failures
where $s' < s$ and $t' < t$.
That is, we see the number of success and failures at a time before
reaching an endpoint. Then the number of subsequent enrollements needed 
to reach an endpoint is SNB($p$, $s-s'$, $t-t'$).
\end{corollary}
\begin{proof}
The result follows directly by realizing that, conditioned on $s'$
success and $t'$ failures, there are $s-s'$ successes needed to reach the
success endpoint and $t-t'$ failures needed to reach the failure endpoint.
\end{proof}

@@

Fig.~\ref{fig:snb_bin_compare} shows the case
where $s=2$, $t=12$, and $p=0.2$.  The probability masses represented in 
black are equal in a) and b) as are the masses in grey.
The probability that $s$
successes are reached in the SNB process is the same as the binomial 
probability of at least two successes. Likewise, the probability that $t$ 
failures are reached in the SNB process is the same as the binomial
probability of zero or one successes.

@@

A similar expression can be derived using the same calculation with
the constraint that $x \leq -\log(p)$. The result
follows from the aforementioned property of the regularized incomplete
beta function.
\end{proof}

@@

The SNB can be used to extend clinical trial methodology
in at least two areas. First, it can be used to construct nested criteria
for early stopping and efficacy. These trials consist of two-stages.
If the failure endpoint is reached in the first stage then the 
trial stops due to futility. Otherwise, the trial enters the second 
stage, which consists of responders from the first stage along with
new enrollees in the second. Thus the design provides
both early testing for futility along with the ability
to make statistically sound decisions sooner than traditional trials.

@@

There are also several avenues for extending the SNB itself. First is the 
development of a stopped negative multinomial distribution allowing
for more than two endpoints. This extension would have clinical trial
applications including dose escalation where the outcomes are response
to a therapy, failure to respond, and an adverse outcome endpoint where
an enrollee is removed from the trial with neither response or non-response
due to side effects.
